Formal Modeling and Verification of a Rate-Monotonic Scheduling Implementation with Real-Time Maude

### 2016-11-02 RMS/final-version/main.tex
finish modifying section III

\subsubsection{Formal Analysis}
Real-Time Maude provides many useful commands and tools to analyze a
given model. For example, \verb|rewrite| allows to execute the model,
symbolically; given an initial state, \verb|search| is used to search
reachable states satisfying the desired properties; the Maude's
\emph{Inductive Theorem Prover} (ITP) can be applied to interactively
prove properties written in \emph{membership equational logic}.

In this paper, we only consider Real-Time Maude's \emph{Linear
  Temporal Logic (LTL) model checker}, which analyzes whether
\emph{each} behavior satisfies a temporal logic formula. \emph{State
  propositions} are defined as terms of sort \verb|Prop|. Their
semantics is defined by conditional equations of the form $\texttt{ceq
} statePattern \texttt{ |= } prop \texttt{ = } b \texttt{ if } cond$,
with $b$ a term of sort \verb|Bool|, stating that $prop$ evaluates to
$b$ in states which are instances of $statePattern$ provided the
condition $cond$ holds. These equations together define $prop$ to hold
in all states $s$ that make $s \texttt{ |= } prop$ evaluate to
\verb|true|. A temporal logic \emph{formula} is constructed by state
propositions and temporal logic operators such as \verb|~|(negation),
\verb|\/|(disjunction), \verb|[]|(``always''),
\verb|U|(``until''). Real-Time Maude supports both \emph{timed} and
\emph{untimed LTL model checking}. The untimed model checking command
\begin{alltt}
  (mc \(s\) |=u \(\Phi\) .)
\end{alltt}
checks whether the temporal logic formula $\Phi$ holds in all
behaviors starting from the initial state $s$, \emph{with no time
  limit}.

\section{The Implementation of RMS}
\label{s:imp}
The implementation written in C is from an industrial avionic control
system.  Interrupts would be triggered by, and only by, the clock
every $T$, which we call \emph{interrupt cycle}. When an interrupt
request occurs, if the system is interruptible, i.e. the interrupt
mask bit is cleared, the handler function $schedule()$ will be
invoked; otherwise $schedule()$ will be pending until the interrupt
mask bit becomes cleared.  The pseudocode of $schedule()$ is shown in
Figure~\ref{a:schedule}, where $taskList$ is the list of periodic
tasks to be scheduled. We assume that the list is in descending order
of priority, and both variables $taskList$ and $timer$ are global. In
this implementation, there is only one kind of interrupt, the period
$T_i$ of each task is a multiple of $T$, and the tasks are
independent, meeting assumption~(A3).

\begin{figure}[!t]
\hrule\vspace{1mm}
  \begin{algorithmic}[1]\small
\Function{$schedule$}{$ $}{}
  \State \Call{$int\_o\!f\!\!f$}{$ $}; \Comment{to disable interrupts} \label{l:1stline}
  \State \Call{$updateStatus$}{$taskList$}; \label{l:updatestatus}
  \State $timer = timer + 1$; \label{l:timer} \label{l:inc}
  \State $p = taskList$;
  \While{$p$} \label{l:startrun1st}
    \If{$p\rightarrow status == \textit{INTERRUPT}$}
      \State \Return; \label{l:return}
    \ElsIf{$p\rightarrow status == \textit{READY}$}      
      \State $p\rightarrow status = \textit{RUNNING}$;
      \State \Call{$int\_on$}{$ $}; \Comment{to enable interrupts} \label{l:endrun1st}
      \State $p\rightarrow function()$; \Comment{to execute the task} \label{l:function}
      \State \Call{$int\_o\!f\!\!f$}{$ $};
      \State $p\rightarrow status = \textit{DORMANT}$;
    \EndIf
    \State $p = p\rightarrow next$;
  \EndWhile
\EndFunction
\Function{$updateStatus$}{$p$}
  \While{$p$}
    \If{$p\rightarrow status == \textit{RUNNING}$} \label{l:startupdate}
      \State $p\rightarrow status = \textit{INTERRUPT}$;
    \EndIf
    \If{$timer~\%~(p\rightarrow period) == 0$} \Comment{the task should be initiated}
      \If{$p\rightarrow status == \textit{DORMANT}$} \Comment{the previous job finishes} 
        \State $p\rightarrow status = \textit{READY}$;
      \Else \Comment{the status is \textit{READY} or \textit{INTERRUPT}}
  \State \Call{$reportTaskError$}{$p$}; \Comment{task misses its deadline}
      \EndIf
    \EndIf \label{l:endupdate}
    \State $p = p\rightarrow next$;
  \EndWhile
\EndFunction
  \end{algorithmic}
\hrule
  \caption{The C-Like Pseudo-code of $schedule()$}
  \label{a:schedule}
\end{figure}

In Figure~\ref{a:schedule}, the handler function $schedule()$ first
updates status of all tasks in $taskList$ via function
$updateStatus()$. This updating actually initiates tasks that should
be scheduled in the current interrupt cycle. Then $schedule()$
traverses the list to execute the ready tasks one by one, or to do a
return when encountering an interrupted\footnote{Note that the status
  \textit{INTERRUPT} indicates the task is interrupted for the moment,
  or was interrupted before but its execution is not complete yet.}
task. As for the function $updateStatus()$, it updates each task in
two steps: firstly, if the task is running, it becomes interrupted;
secondly for the task at its initiation time, if its previous job is
complete, it would be set ready, otherwise it misses its deadline,
producing an error. Notice that $schedule()$ is invoked only when the
interrupt request is handled, not when the interrupt is disabled (the
interrupt mask bit is set). Due to the interrupt mask bit, its
execution cannot be interrupted when it is updating status of tasks or
searching the next task to execute, however, it can be interrupted
while executing some task (Line~\ref{l:function}). This allows the
execution of $schedule()$ to be nested.

For simplicity, in the rest of this paper, we use \emph{scheduling} to
refer to the stage, from the moment when a pending interrupt request
is detected, to the moment when the first should-be-run periodic task
starts executing, i.e. Line~\ref{l:return} or~\ref{l:function} in
Figure~\ref{a:schedule}. Therefore, \emph{scheduling time} mainly
consists of three parts: (i) the time for switching context from the
running task, possibly none, to $schedule()$ when an interrupt request
is handled, (ii) the time spent by $schedule()$ searching and setting
the \emph{first} should-be-run periodic task
(Lines~\ref{l:1stline}-\ref{l:endrun1st} in Figure~\ref{a:schedule}),
and (iii) the time for switching context from $schedule()$ to that
task. \emph{Switching} refers to the stage, from the moment when a
periodic task completes its execution, to the moment when the next
should-be-run periodic task starts executing. \emph{Switching time}
thus also consists of three parts: (i) the time for switching context
from the complete task back to $schedule()$, (ii) the time spent by
$schedule()$ searching and setting the \emph{next} should-be-run
periodic task, and (iii) the time for switching context from
$schedule()$ to that task.

@@

\hide{
At last, to make our model checkable by untimed model checking, it is
reasonable to reset the global variable $timer$ when it reaches an
upper bound while increasing (see Line~\ref{l:timer} in
Figure~\ref{a:schedule}) in the model. Then $timer$ is of sort
\verb|Timer| and the upper bound would be the least common multiple of
the periods of all tasks.}

The global variable $timer$ is reset when it reaches an upper bound
while increasing, which is not shown in detail in
Figure~\ref{a:schedule} but is reasonable. The upper bound is the
least common multiple of the periods of all tasks. A sort \verb|Timer|
is defined to model the $timer$.
\hide{
\begin{verbatim}
  op [_/_] : Nat NzNat -> Timer [ctor] .
\end{verbatim}
}

@@

\subsection{Task Initiation}
\label{ss:init}
Periodic tasks are initiated sequentially by the function
$updateStatus()$ in Figure~\ref{a:schedule}, which is treated as an
instantaneous action in our model. It is modeled by a recursive
function \verb|updateStatus_with_|:
\begin{verbatim}
  op updateStatus_with_ : TaskList Timer 
                            -> TaskList . 
\end{verbatim}
%  eq updateStatus null with TIMER = null .
%  eq updateStatus (TASK :: L) with TIMER
%     = (update TASK with TIMER) 
%       :: (updateStatus L with TIMER) .
which applies function \verb|update_with_| on individual task
sequentially to update the status
(Lines~\ref{l:startupdate}-\ref{l:endupdate} in
Figure~\ref{a:schedule}):
\begin{verbatim}
  op update_with_ : Object Timer ~> Object .
  ceq update < O : PTask | period : T, 
                           status : ST > 
        with TIMER
      = if ST == DORMANT 
        then < O : PTask | status : READY >
        else error fi
      if TIMER rem T == 0 .
  eq update < O : PTask | status : ST > 
       with TIMER
     = if ST == RUNNING 
       then < O : PTask | status : INTERRUPT >
       else < O : PTask |> fi [otherwise] .
\end{verbatim}
with \verb|TIMER| the current value of the global variable $timer$.
Given a task, if \verb|TIMER|($timer$) can be divided by its period
\verb|T|, this task should be initiated.  In the case where the task
should be initiated, it is set \verb|READY| if its \verb|status| is
\verb|DORMANT|; otherwise that means the previous job of this task is
not complete, hence it misses its deadline, producing an
\verb|error|. In the other case where the task should not be
initiated, its \verb|status| changes only if it is \verb|RUNNING|.

We can see that \verb|updateStatus_with_| behaves the same as
$updateStatus()$ in Figure~\ref{a:schedule}.

@@

Unlike periodic tasks, even though the scheduling stage is modeled by
a \verb|Counter|, its functionality is too important to be abandoned.
We divide the behaviors of scheduling into three parts.  The first
part contains its timed behaviors. This part is modeled by regarding
scheduling as a system task of sort \verb|SysTask|. Modeling timed
behaviors of tasks is explained in Section~\ref{ss:timedbehavior}.
The other two parts together define its functionality. The second part
corresponds to Lines~\ref{l:updatestatus}-\ref{l:inc} in
Figure~\ref{a:schedule}. It updates the status of $taskList$ and
increases $timer$ by $1$. This part is modeled by function
\verb|_.startScheduling| which applies instantaneously at the
beginning of scheduling, as shown in rule \verb|interrupt-handle|:
\begin{verbatim}
  op _.startScheduling : System -> System .
  eq (L T STS HW ISRC).startScheduling 
     = ((updateStatus L with T) 
        inc(T) STS HW ISRC) .
\end{verbatim}
The third part corresponds to
Lines~\ref{l:startrun1st}-\ref{l:endrun1st}, searching the first
should-be-run periodic task and setting it to execute. It is modeled
by function \verb|_.finishScheduling|, which applies instantaneously
at the end of scheduling:
\begin{verbatim}
  op _.finishScheduling : System -> System .
  eq (L T STS HW ISRC).finishScheduling
     = (L T (finish scheduling in STS) 
        HW ISRC).run1stTask .
\end{verbatim}
where \verb|finish_in_| resets the counter of task \verb|scheduling|,
and \verb|_.run1stTask| models
Lines~\ref{l:startrun1st}-\ref{l:endrun1st} in
Figure~\ref{a:schedule}, searching the task with highest priority
that has status \verb|INTERRUPT| or \verb|READY| then performing an
\emph{interrupt return} or executing it, respectively.

>>>>

\subsubsection{Formal Analysis}
Real-Time Maude provides many useful commands and tools to analyze a
given model. For example, \verb|rewrite| allows to execute the model,
symbolically; given an initial state, \verb|search| is used to search
reachable states satisfying desired properties; the Maude's
\emph{Inductive Theorem Prover} (ITP) can be applied to interactively
prove properties written in \emph{membership equational logic}.

In this paper, we only consider Real-Time Maude's \emph{Linear
  Temporal Logic (LTL) model checker}, which analyzes whether
\emph{each} behavior satisfies a temporal logic formula. \emph{State
  propositions} are defined as terms of sort \verb|Prop|. Their
semantics is defined by conditional equations of the form $\texttt{ceq
} \mathit{statePattern} \texttt{ |= } \mathit{prop} \texttt{ = } b \texttt{ if } cond$,
with $b$ a term of sort \verb|Bool|, stating that $\mathit{prop}$ evaluates to
$b$ in states which are instances of $\mathit{statePattern}$ provided the
condition $cond$ holds. These equations together define $\mathit{prop}$ to hold
in all states $s$ that make $s \texttt{ |= } \mathit{prop}$ evaluate to
\verb|true|. A temporal logic \emph{formula} is constructed by state
propositions and temporal logic operators such as \verb|~|(negation),
\verb|\/|(disjunction), \verb|[]|(``always''),
\verb|U|(``until''). Real-Time Maude supports both \emph{timed} and
\emph{untimed LTL model checking}. The untimed model checking command
%% \indent
%% $\texttt{(mc } s \texttt{ |=u } \Phi \texttt{ .)}$\\
\begin{alltt}
  (mc \(s\) |=u \(\Phi\) .)
\end{alltt}
checks whether the temporal logic formula $\Phi$ holds in all
behaviors starting from the initial state $s$, \emph{with no time
  limit}.

\section{The Implementation of RMS}
\label{s:imp}
The implementation written in C is from an industrial avionic control
system.  Interrupts would be triggered by, and only by, the clock
every $T$, which we call \emph{interrupt cycle}. When an interrupt
request occurs, if the system is interruptible, i.e. the interrupt
mask bit is cleared, the handler function $\mathit{schedule()}$ will
be invoked; otherwise $\mathit{schedule()}$ will be pending until the
interrupt mask bit becomes cleared.  The pseudocode of
$\mathit{schedule()}$ is shown in Fig.~\ref{f:schedule}, where
$\mathit{taskList}$ is the list of periodic tasks to be scheduled. We
assume that the list is in descending order of priority, and both
variables $\mathit{taskList}$ and $\mathit{timer}$ are global. In this
implementation, there is only one kind of interrupt, the period $T_i$
of each task is a multiple of $T$, and the tasks are independent,
meeting assumption~(A3).

\begin{figure}[!t]
\centering
\includegraphics[width=8.8cm]{rms-pseudocode.pdf}
\caption{The C-Like Pseudo-code of $schedule()$}
\label{f:schedule}
\end{figure}

In Fig.~\ref{f:schedule}, the handler function $\mathit{schedule()}$
first updates status of all tasks in $\mathit{taskList}$ via function
$\mathit{updateStatus()}$. This updating actually initiates tasks that
should be scheduled in the current interrupt cycle. Then
$\mathit{schedule()}$ traverses the list to execute the ready tasks
one by one, or to do a return when encountering an
interrupted\footnote{Note that the status \textit{INTERRUPT} indicates
  the task is interrupted currently, or was interrupted before but its
  execution is not complete yet.}  task. As for the function
$\mathit{updateStatus()}$, it updates each task in two steps: firstly,
if the task is running, it becomes interrupted; secondly for the task
at its initiation time, if its previous job is complete, it would be
set ready, otherwise it misses its deadline, producing an
error. Notice that $\mathit{schedule()}$ is invoked only when the
interrupt request is handled, not when the interrupt is disabled (the
interrupt mask bit is set). Due to the interrupt mask bit, its
execution cannot be interrupted when it is updating status of tasks or
searching the next task to execute, however, it can be interrupted
while executing some task (Line~12). This allows the execution of
$\mathit{schedule()}$ to be nested.

For simplicity, in the rest of this paper, we use \emph{scheduling} to
refer to the stage, from the moment when a pending interrupt request
is detected, to the moment when the first should-be-run periodic task
starts executing, i.e. Line~8 or~12 in
Fig.~\ref{f:schedule}. Therefore, \emph{scheduling time} mainly
consists of three parts: (i)~the time for switching context from the
running task, possibly none, to $\mathit{schedule()}$ when an
interrupt request is handled, (ii)~the time spent by
$\mathit{schedule()}$ searching and setting the \emph{first}
should-be-run periodic task (Lines~2--11 in Fig.~\ref{f:schedule}),
and (iii)~the time for switching context from $\mathit{schedule()}$ to
that task. \emph{Switching} refers to the stage, from the moment when
a periodic task completes its execution, to the moment when the next
should-be-run periodic task starts executing. \emph{Switching time}
thus also consists of three parts: (i)~the time for switching context
from the complete task back to $\mathit{schedule()}$, (ii)~the time
spent by $\mathit{schedule()}$ searching and setting the \emph{next}
should-be-run periodic task, and (iii)~the time for switching context
from $\mathit{schedule()}$ to that task.

@@

\hide{
At last, to make our model checkable by untimed model checking, it is
reasonable to reset the global variable $timer$ when it reaches an
upper bound while increasing (see Line~\ref{l:timer} in
Figure~\ref{f:schedule}) in the model. Then $timer$ is of sort
\verb|Timer| and the upper bound would be the least common multiple of
the periods of all tasks.}

The global variable $timer$ is reset when it reaches an upper bound
while increasing, which is not shown in detail in
Figure~\ref{f:schedule} but is reasonable. The upper bound is the
least common multiple of the periods of all tasks. A sort \verb|Timer|
is defined to model the $timer$.
\hide{
\begin{verbatim}
  op [_/_] : Nat NzNat -> Timer [ctor] .
\end{verbatim}
}

@@

\subsection{Task Initiation}
\label{ss:init}
Periodic tasks are initiated sequentially by the function
$updateStatus()$ in Figure~\ref{f:schedule}, which is treated as an
instantaneous action in our model. It is modeled by a recursive
function \verb|updateStatus_with_|:
\begin{verbatim}
  op updateStatus_with_ : TaskList Timer 
                            -> TaskList . 
\end{verbatim}
%  eq updateStatus null with TIMER = null .
%  eq updateStatus (TASK :: L) with TIMER
%     = (update TASK with TIMER) 
%       :: (updateStatus L with TIMER) .
which applies function \verb|update_with_| on individual task
sequentially to update the status
(Lines~\ref{l:startupdate}-\ref{l:endupdate} in
Figure~\ref{f:schedule}):
\begin{verbatim}
  op update_with_ : Object Timer ~> Object .
  ceq update < O : PTask | period : T, 
                           status : ST > 
        with TIMER
      = if ST == DORMANT 
        then < O : PTask | status : READY >
        else error fi
      if TIMER rem T == 0 .
  eq update < O : PTask | status : ST > 
       with TIMER
     = if ST == RUNNING 
       then < O : PTask | status : INTERRUPT >
       else < O : PTask |> fi [otherwise] .
\end{verbatim}
with \verb|TIMER| the current value of the global variable $timer$.
Given a task, if \verb|TIMER|($timer$) can be divided by its period
\verb|T|, this task should be initiated.  In the case where the task
should be initiated, it is set \verb|READY| if its \verb|status| is
\verb|DORMANT|; otherwise that means the previous job of this task is
not complete, hence it misses its deadline, producing an
\verb|error|. In the other case where the task should not be
initiated, its \verb|status| changes only if it is \verb|RUNNING|.

We can see that \verb|updateStatus_with_| behaves the same as
$updateStatus()$ in Figure~\ref{f:schedule}.

@@

Unlike periodic tasks, even though the scheduling stage is modeled by
a \verb|Counter|, its functionality is too important to be abandoned.
We divide the behaviors of scheduling into three parts.  The first
part contains its timed behaviors. This part is modeled by regarding
scheduling as a system task of sort \verb|SysTask|. Modeling timed
behaviors of tasks is explained in Section~\ref{ss:timedbehavior}.
The other two parts together define its functionality. The second part
corresponds to Lines~\ref{l:updatestatus}-\ref{l:inc} in
Figure~\ref{f:schedule}. It updates the status of $taskList$ and
increases $timer$ by $1$. This part is modeled by function
\verb|_.startScheduling| which applies instantaneously at the
beginning of scheduling, as shown in rule \verb|interrupt-handle|:
\begin{verbatim}
  op _.startScheduling : System -> System .
  eq (L T STS HW ISRC).startScheduling 
     = ((updateStatus L with T) 
        inc(T) STS HW ISRC) .
\end{verbatim}
The third part corresponds to
Lines~\ref{l:startrun1st}-\ref{l:endrun1st}, searching the first
should-be-run periodic task and setting it to execute. It is modeled
by function \verb|_.finishScheduling|, which applies instantaneously
at the end of scheduling:
\begin{verbatim}
  op _.finishScheduling : System -> System .
  eq (L T STS HW ISRC).finishScheduling
     = (L T (finish scheduling in STS) 
        HW ISRC).run1stTask .
\end{verbatim}
where \verb|finish_in_| resets the counter of task \verb|scheduling|,
and \verb|_.run1stTask| models
Lines~\ref{l:startrun1st}-\ref{l:endrun1st} in
Figure~\ref{f:schedule}, searching the task with highest priority
that has status \verb|INTERRUPT| or \verb|READY| then performing an
\emph{interrupt return} or executing it, respectively.
